---
title: "Equality and equivalence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{equality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(k3)
```

Testing R objects for equality is tricky. The `==` operator propagates missing values:

```{r}
NA == 3

NA == NA
```

It treats `NA` as a placeholder for an unknown value. If one or both values are unknown, we can't tell whether or not they are equal. The `NA` result is logically correct. Data are often an incomplete representation of certain real-world facts, and transparency demands that the data reveal this lack of information.

Yet sometimes, our focus is on the data themselves, not on what they may point to in the real world. Rather than asking "Are A and B the same?", we might want to compare *data record* A with *data record* B.

Questions we may ask include:

-   Are A and B equally recorded? In other words: are both `NA`, or else, are both known and equal?

-   Are both A and B missing?

-   Are both A and B known, that is, not missing?

-   Do A and B have the same missingness status? In other words: are both `NA`, or else, are both known?

## Test for equivalence

Use `equivalent()` to test which values evaluate equally:

```{r}
x1 <- c(1, NA, 1, NA)
x2 <- c(1, NA, 2, 5)

equivalent(x1, x2)
```

`1` and `1` are the same value, so they are equivalent. The same is true of `NA` and `NA`. However, `1` and `2` are different, and so are `NA` and `5`. Recall that this function is not concerned with unknown values behind `NA` — it just checks which data are recorded in R.

For convenience, the binary operator `%eq%` can be used instead:

```{r}
x1 %eq% x2
```

If performance is an issue and you only want to compare a single pair of values, use the faster `equivalent_scalar()` instead. It will fail unless both inputs have length 1. The operator `%eq1%` works the same way.

### Going beyond base R

There is nothing like `equivalent()` in base R. `NA == NA` returns `NA` because `==` cares about what these values represent, which is unknown. This is a great way of relating data to the real world. However, `NA` and `NA` are still equal as R objects!

Users tend to call `identical()` instead, which is extremely strict. As soon as two missing values have different types, it will see them as different. Even if both are numbers, `identical()` returns `FALSE` if they are integer (`NA_integer_`) and double (`NA_real_`):

```{r}
# Same type of `NA`
identical(NA, NA)
identical(NA_integer_, NA_integer_)

# Different types of `NA`
identical(NA, NA_integer_)
identical(NA_real_, NA_integer_)
identical(NA_real_, NA_character_)
```

In summary, `==` doesn't acknowledge that two missing values are equal as such, and `identical()` doesn't acknowledge that two missing values of different types are equal at all.

`equivalent()` steers a middle course: by default, it considers any two missing values to be equal. This avoids both extremes, focusing on shared missingness instead. Likewise, one missing and one known value are not equivalent.

### Strict-mode comparison

In base R, double vectors can test equal to integer vectors using `==`. The default for `equivalent()` is the same:

```{r}
x2
typeof(x2)

x3 <- c(3L, NA, 1L, 5L)
typeof(x3)

equivalent(x2, x3)
```

Override this default with `strict = TRUE`:

```{r}
equivalent(x2, x3, strict = TRUE)
```

Here, no values can be equal to begin with, simply because they have different types — even the `NA`s; `strict` simply applies `identical()`. This can be useful if the data type is important but not known in advance.

### Code schema

Below is a schema of the way `equivalent()` compares each pair of values. Note that the actual implementation is more efficient (and iterates over pairs, unlike here).

```{r results='hide'}
function(x, y, strict = FALSE) {
  if (strict) {
    identical(x, y)
  } else if (is.na(x) && is.na(y)) {
    TRUE
  } else if (is.na(x) || is.na(y)) {
    FALSE
  } else {
    x == y
  }
}
```
